# Introduction

```{r introduction_code, echo = FALSE, cache = FALSE, warning=FALSE, message=FALSE}
library(astsa)
library(imudata)
library(simts)

# Function computes direction random walk moves
RW2dimension = function(steps = 100, probs = c(0.25, 0.5, 0.75),
                        couleur = "blue4", xlab = "X-position",
                        ylab = "Y-position", main = NULL, pt_col = NULL,
                        pt_pch = 16, pt.cex = 2, leg_pos = NULL){
  
  # Title
  if (is.null(main)){
    main = paste("Simulated 2D RW with", steps, "steps", sep = " ")
  }
  
  # Points colors
  if (is.null(pt_col)){
    hues = seq(15, 375, length = 3)
    pt_col = hcl(h = hues, l = 65, c = 100)[1:2]
  }
  
  # Initial matrix
  step_direction = matrix(0, steps+1, 2)

  # Start random walk
  for (i in seq(2, steps+1)){
    # Draw a random number from U(0,1)
    rn = runif(1)

    # Go right if rn \in [0,prob[1])
    if (rn < probs[1]) {step_direction[i,1] = 1}

    # Go left if rn \in [probs[1], probs[2])
    if (rn >= probs[1] && rn < probs[2]) {step_direction[i,1] = -1}

    # Go forward if rn \in [probs[2], probs[3])
    if (rn >= probs[2] && rn < probs[3]) {step_direction[i,2] = 1}

    # Go backward if rn \in [probs[3],1]
    if (rn >= probs[3]) {step_direction[i,2] = -1}
  }

  # Cumulative steps
  position = data.frame(x = cumsum(step_direction[, 1]), 
                        y = cumsum(step_direction[, 2]))
  
  par(mar = c(5.1, 5.1, 1, 2.1))

  # Main plot
  plot(NA, xlim = range(position[,1]), ylim = range(range(position[,2])), 
       xlab = xlab, ylab = ylab, xaxt = 'n', 
       yaxt = 'n', bty = "n", ann = FALSE)
  win_dim = par("usr")

  par(new = TRUE)
  plot(NA, xlim = range(position[,1]), ylim = c(win_dim[3], win_dim[4] + 0.09*(win_dim[4] - win_dim[3])),
       xlab = xlab, ylab = ylab, xaxt = 'n', yaxt = 'n', bty = "n")
  win_dim = par("usr")

  # Add grid
  grid(NULL, NULL, lty = 1, col = "grey95")

  # Add title
  x_vec = c(win_dim[1], win_dim[2], win_dim[2], win_dim[1])
  y_vec = c(win_dim[4], win_dim[4],
            win_dim[4] - 0.09*(win_dim[4] - win_dim[3]),
            win_dim[4] - 0.09*(win_dim[4] - win_dim[3]))
  polygon(x_vec, y_vec, col = "grey95", border = NA)
  text(x = mean(c(win_dim[1], win_dim[2])), y = (win_dim[4] - 0.09/2*(win_dim[4] - win_dim[3])), main)

  # Add axes and box
  lines(x_vec[1:2], rep((win_dim[4] - 0.09*(win_dim[4] - win_dim[3])),2), col = 1)
  box()
  axis(1, padj = 0.3)
  y_axis = axis(2, labels = FALSE, tick = FALSE)
  y_axis = y_axis[y_axis < (win_dim[4] - 0.09*(win_dim[4] - win_dim[3]))]
  axis(2, padj = -0.2, at = y_axis)

  # Add trajectory
  lines(position, type = "l", col = couleur, pch = 16)
  
  # Start and end points
  points(c(0,position[steps+1,1]), c(0,position[steps+1,2]), cex = pt.cex,
         col = pt_col, pch = pt_pch)
  
  # Legend
  if (is.null(leg_pos)){
    leg_pos = c(min(position[,1]), max(position[,2]))
  }
  legend(leg_pos[1], leg_pos[2], c("Start","End"), 
         col = pt_col, pch = pt_pch, pt.cex = pt.cex, bty = "n")
}
```

<!--
> "One damned thing after another." ~ R. A. Fisher
>
-->

>
> "*Prévoir consiste à projeter dans l’avenir ce qu’on a perçu dans le passé.*" – Henri Bergson
>

<!--
>
> *Hâtons-nous; le temps fuit, et nous entraîne avec soi : le moment où je parle est déjà loin de moi*. Nicolas Boileau
>
-->

After reading this chapter you will be able to:

- Describe what a *time series* is.
- Perform exploratory data analysis on time series data.
- Evaluate different characteristics of a time series.
- Classify basic time series models through equations and plots. 
<!-- - Manipulate a time series equation using *backsubstitution*. -->

## Time Series

A *time series* is a particular kind of *stochastic process* which, generally speaking, is a collection of random variables indexed by a set of numbers. Not surprisingly, the index of reference for a time series is given by *time* and, consequently, a time series is a collection of random variables indexed (or "measured") over time such as, for example, the daily price of a financial asset or the monthly average temperature in a given location. In terms of notation, a time series is often represented as

 \[\left(X_1, X_2, ..., X_n \right) \;\;\; \text{ or } \;\;\; \left(X_t\right)_{t = 1,...,n}.\]
 
The time index $t$ is contained within either the set of reals, $\mathbb{R}$, or
integers, $\mathbb{N}$. When $t \in \mathbb{R}$, the time series becomes a
*continuous-time* stochastic process such a Brownian motion, a model used to
represent the random movement of particles within a suspended liquid or gas,
or an ElectroCardioGram (ECG) signal, which corresponds to the palpitations of
the heart. However, within this text, we will limit ourselves to the cases where
$t \in \mathbb{N}$, better known as *discrete-time* processes.
*Discrete-time* processes are where a variable is measured sequentially at fixed
and equally spaced intervals in time.
This implies that we will have two assumptions:

1. $t$ is not random, e.g. the time at which each observation is measured is known, and
2. the time between two consecutive observations is constant. 

```{r numberline, echo = FALSE, cache = TRUE, fig.height = 1.5, fig.align = 'center', fig.cap = "Discrete-time can be thought of as viewing a number line with equally spaced points."}
```

Moreover, the term "time series" can also represent a probability model for a 
set of observations. For example, one of the fundamental probability models used
in time series analysis is called a *white noise* process and is defined as

\[W_t \mathop \sim \limits^{iid} N(0, \sigma^2).\]

This statement simply means that $(W_t)$ is normally distributed and independent
over time. This model may appear to be dull but we will soon see it as a crucial
component to constructing more complex models. Unlike the white noise process,
time series are typically *not* independent over time. Suppose that the 
temperature in Champaign is unusually low, then it is reasonable to assume
that tomorrow's temperature will also be low. Indeed, such behavior would
suggest the existence of a dependency over time. The time series methods we will
discuss in this text consists of parametric models used to characterize
(or at least approximate) the joint distribution of $(X_t)$. Often, time series
models can be decomposed into two components, the first of which is what we call
a *signal*, say $(Y_t)$, and the second component is a *noise*, say $(W_t)$,
leading to the model

 \[X_t = Y_t + W_t.\]
 
Typically, we have $\mathbb{E}[Y_t] \neq 0$ while $\mathbb{E}[W_t] = 0$ (although we may have
$\mathbb{E}[W_t | W_{t-1}, ..., W_1] \neq 0$). Such models impose some parametric
structure which represents a convenient and flexible way of studying time series
as well as a means to evaluate *future* values of the series through forecasting.
As we will see, predicting future values is one of the main aspects of time
series analysis. However, making predictions is often a daunting task or as
famously stated by Nils Bohr:

> 
> "*Prediction is very difficult, especially about the future.*"
>

There are plenty of examples of predictions that turned out to be completely
erroneous. For example, three days before the 1929 crash, Irving Fisher,
Professor of Economics at Yale University, famously predicted:

>
> "*Stock prices have reached what looks like a permanently high plateau*". 
>

Another example is given by Thomas Watson, president of IBM, who said in 1943:

>
> "*I think there is a world market for maybe five computers.*"
>

## Exploratory Data Analysis for Time Series {#eda}

When dealing with relatively small time series (e.g. a few thousands), it is
often useful to look at a graph of the original data. A graph can be
an informative tool for "detecting" some features of a time series such as trends and
the presence of outliers.

Indeed, a trend is typically assumed to be present in a time series when the
data exhibit some form of long term increase or decrease or combination of
increases or decreases. Such trends could be linear or non-linear and represent
an important part of the "signal" of a model. Here are a few examples of
non-linear trends:

1. **Seasonal trends** (periodic): These are the cyclical patterns which repeat
after a fixed/regular time period. This could be due to business cycles
(e.g. bust/recession, recovery).
    
2. **Non-seasonal trends** (periodic): These patterns cannot be associated to
seasonal variation and can for example be due to an external variable such as,
for example, the impact of economic indicators on stock returns. Note that such
trends are often hard to detect based on a graphical analysis of the data.
    
3. **"Other" trends**: These trends have typically no regular patterns and are
over a segment of time, known as a "window", that change the statistical
properties of a time series. A common example of such trends is given by the
vibrations observed before, during and after an earthquake.

```{example, label="jjquarterly"}
A traditional example of a time series is the quarterly earnings of
the company Johnson and Johson. In the figure below, we present these earnings
between 1960 and 1980.
```

```{r example_JJ, fig.height = 4, fig.width = 7, cache = TRUE}
# Load data
data(jj, package = "astsa")

# Construct gts object
jj = gts(jj, start = 1960, freq = 4, unit_ts = "$", name_ts = "Earnings", 
         data_name = "Johnson and Johnson Quarterly Earnings")

# Plot time series
plot(jj)
```

One trait that the graph makes evident is that the data contains a non-linear
increasing trend as well as a yearly seasonal component. In addition, one can
note that the *variability* of the data seems to increase with time. Being able
to make such observations provides important information to select suitable
models for the data.

Moreover, when observing "raw" time series data it is also interesting to
evaluate if some of the following phenomena occur:

1. **Change in Mean:** Does the mean of the process shift over time?
2. **Change in Variance:** Does the variance of the process evolve with time?
3. **Change in State:** Does the time series appear to change between "states"
   having distinct statistical properties?
4. **Outliers** Does the time series contain some "extreme" observations? 
   (Note that this is typically difficult to assess visually.)

```{example, label="earthquake"}
In the figure below, we present an example of displacement recorded
during an earthquake as well as an explosion.
```

```{r example_EQ, fig.height = 4, fig.width = 7, cache = TRUE}
# TO DO!
```

From the graph, it can be observed that the statistical properties of the time
series appear to change over time. For instance, the variance of the time series
shifts at around $t = 1150$ for both series.
The shift in variance also opens "windows" where there appear to be distinct
states. In the case of the explosion data, this is particularly relevant around
$t = 50, \cdots, 250$ and then again from $t = 1200, \cdots, 1500$. Even within
these windows, there are "spikes" that could be considered as outliers most
notably around $t = 1200$ in the explosion series.

Extreme observations or outliers are commonly observed in real time series data, this is illustrated in the following example. 

```{example, label="percipitation"}
We consider here a data set coming from the domain of hydrology. The data 
concerns monthly precipitation (in mm) over a certain period of time (1907 to
1972) and is interesting for scientists in order to study water cycles. The
data are presented in the graph below:
```

```{r example_hydro, fig.height = 4, fig.width = 7, cache = TRUE, message = FALSE, warning = FALSE}
# Load hydro dataset
data("hydro")

# Simulate based on data
hydro = gts(as.vector(hydro), start = 1907, freq = 12, unit_ts = "in.", 
            name_ts = "Precipitation", data_name = "Hydrology data")

# Plot hydro 
plot(hydro)
```

Next, we consider an example coming from high-frequency finance to illustrate
the limitations our current framework.

```{example, label="stocks-starbucks"}
The figure below presents the returns or price innovations
(i.e. the changes in price from one observation to the
next) for Starbuck's stock on July 1, 2011 for about 150 seconds (left
panel) and about 400 minutes (right panel). 
```

```{r example_Starbucks, fig.height = 4, fig.width = 7, cache = TRUE, message = FALSE}
# TO DO add Duke's code here!
```

It can be observed on the left panel that observations are not equally spaced.
Indeed, in high-frequency data the intervals between two points are typically not
constant and are, even worse, random variables. This implies that the time when
a new observation will be available is in general unknown. On the right panel,
one can observe that the variability of the data seems to change during the
course of the trading day. Such a phenomenon is well known in the finance
community since a lot of variation occurs at the start (and the end) of the 
day while the middle of the day is associated with small changes. 
Moreover, clear extreme observations can also be noted in this graph at
around 11:00.

Finally, let us consider the limitations of a direct graphical representation of
a time series when the sample size is large. Indeed, due to visual limitations,
a direct plotting of the data will probably result in an uninformative
aggregation of points between which it is unable to distinguish anything. This is
illustrated in the following example.

```{example, label="large-imu"}
We consider here the data coming from the calibration procedure of
an Inertial Measurement Unit (IMU) which, in general terms, is used to enhance
navigation precision or reconstruct three dimensional movements (see
e.g. [link](https://www.youtube.com/watch?v=htoBvSq8jLA)). These sensors are
used in a very wide range of applications such as robotics, virtual reality, 
vehicle stability control, human and animal motion capture and so forth (see e.g. [link](https://www.youtube.com/watch?v=g4tgtPA54_Y)).
The signals coming from these instruments are measured at
high frequencies over a long time and are often characterized by linear trends
and numerous underlying stochastic processes.
```
The code below retrieves some data from an IMU and plots it directly:

```{r example_IMU, fig.height = 4, fig.width = 7, cache = TRUE, message = FALSE, dev='png'}
# Load IMU data
data(imu6, package = "imudata")

# Construct gst object
Xt = gts(imu6[,1], data_name = "Gyroscope data", unit_time = "hour", 
         freq = 100*60*60, name_ts = "Angular rate", 
         unit_ts = bquote(rad^2/s^2))

# Plot time series
plot(Xt)
```

Although a linear trend and other processes are present in this signal
(time series), it is practically impossible to understand or guess anything
from the plot.

## Basic Time Series Models {#basicmodels}

In this section, we introduce some simple time series models. Before doing so
it is useful to define $\Omega_t$ as all the information avaiable up to time
$t-1$, i.e.

\[\Omega_t = \left(X_{t-1}, X_{t-2}, ..., X_0 \right).\]

As we will see this compact notation is quite useful.

### White noise processes {#wn}

The building block for most time series models is the Gaussian white noise
process, which can be defined as

\[{W_t}\mathop \sim \limits^{iid} N\left( {0,\sigma _w^2} \right).\]

This definition implies that:

1. $\mathbb{E}[W_t | \Omega_t] = 0$ for all $t$,
2. $\text{cov}\left(W_t, W_{t-h} \right) = \boldsymbol{1}_{h = 0} \; \sigma^2$ for
all $t, h$.

Therefore, in this process there is an absence of temporal (or serial) 
dependence and is homoskedastic (i.e. it has a constant variance). 
White noise can be generalized into two sorts of processes: *weak* and *strong*. The process $(W_t)$ is
a weak white noise if

1. $\mathbb{E}[W_t] = 0$ for all $t$,
2. $\text{var}\left(W_t\right) = \sigma^2$ for all $t$,
3. $\text{cov} \left(W_t, W_{t-h}\right) = 0$, for all $t$, and for all $h \neq 0$.

Note that this definition does not imply that $W_t$ and $W_{t-h}$ are
independent (for $h \neq 0$) but simply uncorrelated.
However, the notion of independence is used to define a *strong* white noise as

1. $\mathbb{E}[W_t] = 0$ and $\text{var}(W_t) = \sigma^2 < \infty$, for all $t$,
2. $F(W_t) = F(W_{t-h})$, for all $t,h$ (where $F(W_t)$ denotes the distribution of $W_t$),
3. $W_t$ and $W_{t-h}$ are independent for all $t$ and for all $h \neq 0$.

It is clear from these definitions that if a process is a strong white noise
it is also a weak white noise. However, the converse is not true as shown in
the following example:

```{example, label="weaknotstrong"}

Let $Y_t \mathop \sim F_{t+2}$, where $F_{t+2}$ denotes
a Student distribution with $t+2$ degrees of freedom. Assuming the 
sequence $(Y_1, \ldots, Y_n)$ to be independent, we 
let $X_t = \sqrt{\frac{t}{t+2}} Y_t$. Then, the process $(X_t)$ is obviously
not a strong white noise as the distribution of $X_t$ changes with $t$. However,
this process is a weak white noise since we have:

- $\mathbb{E}[X_t] = \sqrt{\frac{t}{t+2}} \mathbb{E}[Y_t] = 0$ for all $t$.
- $\text{var}(X_t) = \frac{t}{t+2} \text{var}(Y_t) = \frac{t}{t+2} \frac{t+2}{t} = 1$ for all $t$.
- $\text{cov}(X_t, X_{t+h}) = 0$ (by independence), for all $t$, and for all $h \neq 0$.

```

The code below presents an example of how to simulate a Gaussian white noise process.

```{r example_WN, fig.height = 4, fig.width = 7, cache = TRUE}
n = 1000                               # process length
sigma2 = 1                             # process variance
Xt = gen_gts(n, WN(sigma2 = sigma2))
plot(Xt)
```

### Random Walk Processes {#rw}

The term *random walk* was first introduced by Karl Pearson in the early
nineteen-hundreds. Regarding white noise, there exist a large range of random walk
processes. For example, one of the simplest forms of a random walk process can be
explained as follows: suppose that you are walking on campus and your
next step can either be to your left, your right, forward or backward
(each with equal probability). Two realizations of such processes are
represented below:

```{r RW2d, fig.height = 5, fig.width = 5.5, cache = TRUE, fig.align='center'}
set.seed(5)
RW2dimension(steps = 10^2)
RW2dimension(steps = 10^4)
```

Such processes inspired Karl Pearson's famous quote that

>
> "*the most likely place to find a drunken walker is somewhere near his starting point.*"
> 

Empirical evidence of this phenomenon is not too hard to find on a Friday night. In this text, we only consider one very specific form of
random walk, namely the Gaussian random walk which can be defined as:

$$X_t = X_{t-1} + W_t,$$

where $W_t$ is a Gaussian white noise process with initial condition $X_0 = c$.
(Typically $c = 0$.) This process can be expressed differently by
*backsubstitution* as follows:

\[\begin{aligned}
  {X_t} &= {X_{t - 1}} + {W_t} \\
   &= \left( {{X_{t - 2}} + {W_{t - 1}}} \right) + {W_t} \\
   &= \vdots \\
  {X_t} &= \sum\limits_{i = 1}^t {{W_i}} + X_0 =  \sum\limits_{i = 1}^t {{W_i}} + c \\ 
\end{aligned} \]

The code below presents an example of how to simulate a such process.

```{r example_RW, fig.height = 4, fig.width = 7, cache=TRUE}
n = 1000                               # process length
gamma2 = 1                             # innovation variance
Xt = gen_gts(n, RW(gamma2 = gamma2))
plot(Xt)
```



### Autoregressive Process of Order 1 {#ar1}

An autoregressive process of order 1 or AR(1) is a generalization of both
the white noise and random walk processes which are both themselves special
cases of an AR(1). A (Gaussian) AR(1) process can be defined as

\[{X_t} = {\phi}{X_{t - 1}} + {W_t},\]

where $W_t$ is a Gaussian white noise. Clearly, an AR(1) with $\phi = 0$ is
a Gaussian white noise and when $\phi = 1$ the process becomes a random walk. 

```{remark, label="ar1mean"}
We generally assume that an AR(1), as well as other time series
models, have zero mean. The reason for this assumption is only to simplfy the
notation but it is easy to consider an AR(1) process around an
arbitrary mean $\mu$, i.e.

\[\left(X_t - \mu\right) = \phi \left(X_{t-1} - \mu \right) + W_t,\]

which is of course equivalent to

\[X_t = \left(1 - \phi \right) \mu + \phi X_{t-1} + W_t.\]

Thus, we will generally only work with zero mean processes since adding means
is simple.
```

```{remark, label="ar1realizations"}
An AR(1) is in fact a linear combination of past realisations of
the white noise $W_t$ process. Indeed, we have

\[\begin{aligned}
 {X_t} &= {\phi_t}{X_{t - 1}} + {W_t} 
   = {\phi}\left( {{\phi}{X_{t - 2}} + {W_{t - 1}}} \right) + {W_t} \\
   &= \phi^2{X_{t - 2}} + {\phi}{W_{t - 1}} + {W_t} 
   = {\phi^t}{X_0} + \sum\limits_{i = 0}^{t - 1} {\phi^i{W_{t - i}}}.
\end{aligned}\]

Under the assumption of infinite past (i.e. $t \in \mathbb{Z}$) and $|\phi| < 1$,
we obtain

\[X_t = \sum\limits_{i = 0}^{\infty} {\phi^i {W_{t - i}}},\]

since $\operatorname{lim}_{i \to \infty} \; {\phi^i}{X_{t-i}} = 0$.
```

The code below presents an example of how an AR(1) can be simulated.

```{r example_AR1, fig.height = 4, fig.width = 7, cache=TRUE}
n = 1000                              # process length
phi = 0.5                             # phi parameter
sigma2 = 1                            # innovation variance
Xt = gen_gts(n, AR1(phi = phi, sigma2 = sigma2))
plot(Xt)
```

### Moving Average Process of Order 1 {#ma1}

As we have seen in the previous example, an AR(1) can be expressed as a linear
combination of all past observations of $(W_t)$ while the next process, called
a moving average process of order 1 or MA(1), is (in some sense) a "truncated"
version of an AR(1). It is defined as

\begin{equation} 
  X_t = \theta W_{t-1} + W_t,
\end{equation}
 
where (again) $W_t$ denotes a Gaussian white noise process. An example on how to
generate an MA(1) is given below:

```{r example_MA1, fig.height = 4, fig.width = 7, cache=TRUE}
n = 1000                              # process length
sigma2 = 1                            # innovation variance
theta = 0.5                           # theta parameter
Xt = gen_gts(n, MA1(theta = theta, sigma2 = sigma2))
plot(Xt)
```

  
### Linear Drift {#drift}

A linear drift is a very simple deterministic time series model which can be
expressed as 

\[X_t = X_{t-1} + \omega, \]

where $\omega$ is a constant and with the initial condition $X_0 = c$, where $c$ is an
arbitrary constant (typically zero). This process can be expressed in a more
familiar form as follows:

\[
  {X_t} = {X_{t - 1}} + \omega 
   = \left( {{X_{t - 2}} + \omega} \right) + \omega 
   = t{\omega} + c  \]

Therefore, a (linear) drift corresponds to a simple linear model with slope
$\omega$ and intercept $c$.

A drift can simply be generated using the code below:

```{r example_Drift, fig.height = 4, fig.width = 7, cache=TRUE}
n = 100                               # process length
omega = 0.5                           # slope parameter
Xt = gen_gts(n, DR(omega = omega))
plot(Xt)
```


## Composite Stochastic Processes {#lts}

A composite stochastic process can be defined as the sum of underlying 
(or latent) stochastic processes. In this text, we will use the term
*latent time series* as a synomym for composite stochastic processes.
A simple example of such a process is given by

\[\begin{aligned}
Y_t &= Y_{t-1} + W_t + \delta\\
X_t &= Y_t + Z_t,
\end{aligned}\]

where $W_t$ and $Z_t$ are two independent Gaussian white noise processes. 
This model is often used as a first tool to approximate the number of
individuals in the context ecological population dynamics. 
For example, suppose we want to study the population of Chamois in the Swiss Alps.
Let $Y_t$ denote the "true" number of individuals in this population at time $t$.
It is reasonable that $Y_t$ is (approximately) the population at the previous
time $t-1$ (e.g the previous year) plus a random variation and a drift. 
This random variation is due to the natural randomness in ecological population
dynamics and reflects changes such as the number of predators, the abundance
of food, or weather conditions.
On the other hand, ecological *drift* is often of particular interest for ecologists as 
it can be used to determine the "long" term trends of the population 
(e.g. if the population is increasing, decreasing, or stable).
Of course, $Y_t$ (the number of individauls) is typically unknown and we observe
a noisy version of it, denoted as $X_t$. 
This process corresponds to the true population plus a measurement error since
some individuals may not be observed while others may have been counted several
times.
Interestingly, this process can clearly be expressed as a
*latent time series model* (or composite stochastic process) as follows:

\[\begin{aligned}
R_t &= R_{t-1} + W_t \\
S_t &= \delta t \\
X_t &= R_t + S_t + Z_t,
\end{aligned}\]

where $R_t$, $S_t$ and $Z_t$ denote, respectively, a random walk,
a drift, and a white noise. The code below can be used to simulate such data:

```{r example_composite, fig.height = 6, fig.width = 7, cache=TRUE}
n = 1000                                # process length
delta = 0.005                           # delta parameter (drift)
sigma2 = 10                             # variance parameter (white noise)
gamma2 = 0.1                            # innovation variance (random walk)
model = WN(sigma2 = sigma2) + RW(gamma2 = gamma2) + DR(omega = delta)
Xt = gen_lts(n, model)
plot(Xt)
```

In the above graph, the first three plots represent the latent (unobserved)
processes (i.e. white noise, random walk, and drift) and the last one represents
the sum of the three (i.e. $(X_t)$).

Let us consider a real example where these latent processes are useful to
describe (and predict) the behavior of economic variables such as Personal 
Saving Rates (PSR). A process that is used for these settings is the 
"random-walk-plus-noise" model, meaning that the data can be explained by a 
random walk process in addition to which we observe some other process (e.g. 
a white noise model, an autoregressive model such as an AR(1), etc.). The PSR 
taken from the Federal Reserve of St. Louis from January 1, 1959, to  May 1,
2015, is presented in the following plot:

```{r example_PSR, fig.height = 4, fig.width = 7, cache=TRUE}
# Load savingrt dataset
data("savingrt")
# Simulate based on data
savingrt = gts(as.vector(savingrt), start = 1959, freq = 12, unit_ts = "%", 
            name_ts = "Saving Rates", data_name = "US Personal Saving Rates")
# Plot savingrt simulation
plot(savingrt)
```

It can be observed that the mean of this process seems to vary over time,
suggesting that a random walk can indeed be considered as a possible model
to explain this data. In addition, aside from some "spikes" and occasional 
sudden changes, the observations appear to gradually change from one time point
to the other, suggesting that some other form of dependence between them could
exist.
